---
title: "Lab 2 - Data wrangling"
author: "Jakob Bickley"
date: "Feb 4, 2026"
format: 
  html: default # for speed
  pdf: default # for final output
---

```{r}
#| eval: true
#| message: false

library(tidyverse)
```

# Questions

## Part 1

### Question 1

```{r}
midwest <- midwest
midwest |> group_by(state)|> summarize(n = n()) |> arrange(desc(n))
```

Illinois has the greatest number of counties at 102, and Wisconsin has the lowest at 72, across Midwest states.

\newpage

### Question 2

```{r}
midwest_duplicate <- midwest |> group_by(county) |> summarize(count = n()) |> filter(count >= 2) |> arrange(desc(count))
midwest_duplicate
```

\newpage

### Question 3

```{r}
midwest |> filter(popdensity > 25000) |> select(county, state, popdensity, poptotal, area) |> arrange(desc(popdensity))
midwest |> select(county, state, popdensity, poptotal, area) |> filter(popdensity == max(popdensity))
```

\newpage

### Question 4

The distribution of population density of counties is unimodal and extremely right-skewed. A typical Midwestern county has population density of `r summarise(midwest, median_density = median(popdensity))` people per unit area. The middle 50% of the counties have population densities between `r summarise(midwest, lower = quantile(popdensity, .25))` to `r summarise(midwest, lower = quantile(popdensity, .75))` people per unit area.
\# Numbers are inline code 
\# Used AI to get quantile() function

\newpage

### Question 5

\newpage

### Question 6

\newpage

### Question 7

\newpage

### Question 8

\newpage

## Part 2

### Question 9
